Inferring probabilistic stellar rotation periods using Gaussian processes

### 2017-02-02 documents/ms.tex
tdm abstract edits

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior Probability Density Function (PDF)
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We use \naigrain\ simulated light curves with a range of rotation periods to
    test the GP model and compare our results with a sine-fitting periodogram
    method and an AutoCorrelation Function (ACF) method.
The posterior PDFs of the parameters were sampled using the affine invariant
    ensemble Markov Chain Monte Carlo (MCMC) sampler {\tt emcee3}
    \citep{Foreman-Mackey2013}, and the GP operations were performed using the
    {\tt george} python package \citep{George}.
We demonstrate that rotation periods inferred via this method are more
    accurate than the periodogram and ACF methods.
Furthermore, the rotation periods are inferred probabilistically, allowing
    them to be easily incorporated into follow-on studies; inferring stellar
    ages or characterising star-planet interactions for example.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

>>>>

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes, causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid therefore cannot accurately model a 
rotationally modulated stellar light curve.
Physical models of the stellar surface also have many drawbacks preventing effective
inference, such as having highly degenerate or high-dimensional parameter spaces
\citep[\eg][]{Russell1906, Jeffers2009, Kipping2012}.
In this work, we introduce an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
Using this highly flexible model, 
we are able to sample from the posterior Probability Density Function
(PDF) of the periodic parameter, marginalising over the other kernel hyperparameters
using a Markov Chain Monte Carlo approach.
To test this method's effectiveness, we infer rotation periods from 
\naigrain\ simulated light curves, demonstrating that our results 
    are more accurate than both a sine-fitting periodogram and an 
    AutoCorrelation Function (ACF) method.
We also demonstrate that it works effectively on real data, by 
inferring rotation periods for \nkois\ \Kepler\ stars.
Because this method delivers posteriors PDFs, 
it will enable other studies involving stellar rotation, 
such as inferring stellar
ages or stellar obliquities in exoplanet systems, or characterising star-planet interactions.
% Thehe ACF method is not applicable to unevenly spaced data,
% therefore it is inappropriate for the ~$10^7$ stellar light curves expected
% to be produced by \LSST\ unlike the GP method.
% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.
